Estimating demographic bias on tests of children's early vocabulary

### 2022-05-27 paper/cogsci_final.Rmd
minor edit
final render

Demographic differences in language ability are likely to be at least partially explained by differences in linguistic input received by children in different groups.
Indeed, input quantity (total daily tokens) and some measures of quality (e.g., lexical diversity: ratio of word types vs. tokens) have often been predictive of language learning outcomes in  demographic studies [@rowe2009differences; @huttenlocher1991early].
We used a corpus-based analysis to investigate the extent to which word frequency in child-directed speech to male vs. female children was predictive of the amount of sex-based bias shown by CDI items. 
Similar to the approach taken by @braginsky2016gender, we used the CHILDES corpus of transcripts from dyadic play sessions [@macwhinney2000childes], which are labeled with the sex of the target child, but not other demographic variables.
We found a total of 5213750 female-directed tokens and 6091950 male-directed tokens that matched 662 of the 680 CDI:WS items, normalized the word frequencies to tokens per million for each target sex, and calculated the log-odds of each word being spoken to a female (i.e., $log(p_{f} / p_{m}$), meaning unbiased words will have log-odds of 0, while those spoken more often to females will have log-odds $>0$, and those spoken more often males will have log-odds $<0$.
For example "doll" was spoken 156 times to females, and 138 times to males, thus $p_{f} = 156/(156+138) = .53$, $p_{m} = 1-p_{f} = 0.47$, and $log(p_{f}/p_{m})$ = 0.12), while "police" was spoken 138 times to females, and 225 times to males, and thus has log-odds = -0.49).
Overall, the correlation between the log-odds of a word being spoken to a female vs. a male child and the size of the female (vs. male) advantage for that CDI word was weak, but significant ($r=0.18$, $p<.001$). 
Some of the sex bias seen in CDI items is related to differences in language input received by girls vs. boys -- though of course this differential input could be elicited due to different interests.

@@

We thank members of the Language and Cognition lab and three anonymous reviewers for their feedback.

>>>>

Demographic differences in language ability are likely to be at least partially explained by differences in linguistic input received by children in different groups.
Indeed, input quantity (total daily tokens) and some measures of quality (e.g., lexical diversity: ratio of word types vs. tokens) have often been predictive of language learning outcomes in  demographic studies [@rowe2009differences; @huttenlocher1991early].
In an attempt to investigate this, we analyzed the extent to which word frequency in a corpus of child-directed speech to male vs. female children was predictive of the amount of sex-based bias shown by CDI items. 
Similar to the approach taken by @braginsky2016gender, we used the CHILDES corpus of transcripts from dyadic play sessions [@macwhinney2000childes], which are labeled with the sex of the target child, but not other demographic variables.
We found a total of 5213750 female-directed tokens and 6091950 male-directed tokens that matched 662 of the 680 CDI:WS items, normalized the word frequencies to tokens per million for each target sex, and calculated the log-odds of each word being spoken to a female (i.e., $log(p_{f} / p_{m}$), meaning unbiased words will have log-odds of 0, while those spoken more often to females will have log-odds $>0$, and those spoken more often males will have log-odds $<0$.
For example "doll" was spoken 156 times to females, and 138 times to males, thus $p_{f} = 156/(156+138) = .53$, $p_{m} = 1-p_{f} = 0.47$, and $log(p_{f}/p_{m})$ = 0.12), while "police" was spoken 138 times to females, and 225 times to males, and thus has log-odds = -0.49).
Overall, the correlation between the log-odds of a word being spoken to a female vs. a male child and the size of the female (vs. male) advantage for that CDI word was weak, but significant ($r=0.18$, $p<.001$). 
Some of the sex bias seen in CDI items is related to differences in language input received by girls vs. boys -- though of course this differential input could be elicited due to different interests.

@@

This research was supported by the Stanford Maternal and Child Health Research Institute. We thank members of the Language and Cognition lab and three anonymous reviewers for their feedback.
